Handling Recursion in Generic Programming Using Closed Type Families

### 2018-04-26 lncs-paper.tex
Remove appendices

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to view a datatype as its isomorphic code whose kind is a list of lists of types. Using the \textsf{DataKinds} extension that enables datatype definitions to be promoted to kinds, SOP expresses this code with a type family:
\texttt{
\vs
\indent\textbf{type family} Code (a \ann *) \ann [[*]]
\vs
}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments depending on the chosen constructor; the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.

For example, consider a datatype of binary trees:
\texttt{
\vs
\indent\textbf{data} Tree a = Leaf a | Node (Tree a) (Tree a)
\vs
}
This datatype is isomorphic to the following code:
\texttt{
\vs
\indent\textbf{type instance} Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\vspace{0.2cm}
}

@@

A type of promoted lists has no inhabitants, so the universe provides datatypes to operate on $n$-ary sums and products as on terms. As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types. The universe defines these with embedded functor application, where the functor is some type constructor. Thus, each component of \texttt{NS} and \texttt{NP} corresponds to one type (its index) in the type-level list and has type \texttt{f x}, for an index type \texttt{x}. The definitions of \texttt{NS} and \texttt{NP} both are kind polymorphic. The index list allows to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.

As basic instantiation types of \texttt{f}, SOP defines an identity functor \texttt{I} that is type-level equivalent for \texttt{id}, and a constant functor \texttt{K}, for \texttt{const}:
\texttt{
\vs
\indent\textbf{newtype} I (a \ann *)\ind\ind\s\s\nhs = I \{unI \ann a\}\\
\indent\textbf{newtype} K (a \ann *) (b \ann k) = K \{unK \ann a\}
\vs
}
Whereas instantiated with \texttt{I}, \texttt{NP} is a direct heterogeneous list of types, with \texttt{K a}, it is essentially homogeneous. An example value of an \texttt{NP} for \texttt{I} looks thus:
\texttt{
\vs
\indent I 5 :* I True :* I 'x' :* Nil \ann NP I '[Int, Bool, Char]
\vspace{0.2cm}
}

A clear idea underlies the sum definition. The constructor \texttt{S} of a sum skips the first element of an $n$-element index list giving an index in a list that has $n+1$ elements, while \texttt{Z} stores the payload of type \texttt{f x}. For example, the following chooses the third element of a sum:
\texttt{
\vs
\indent S (S (Z (I 3))) \ann NS I '[Char, Bool, Int, Bool]
\vs
}
The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and gives the product representing arguments of that constructor.

With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism between a datatype and its generic representation:
\texttt{
\vs
\indent\textbf{type} Rep a = SOP I (Code a)
\vs
\indent\textbf{class} All SListI (Code a) \Ra Generic (a \ann *) \textbf{where}\\
\indent\s\s \textbf{type} Code a \ann [[*]]\\
\indent\s\s from \ann a \ra Rep a\\
\indent\s\s to\s\s\s \ann Rep a \ra a
\vs
}
The sum of products \texttt{SOP f} is a newtype for \texttt{NS (NP f)}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code. The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.

The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be satisfied, and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the constraint \texttt{All} specifying that a particular constraint holds for each member of a list of types\footnote{The definition of \texttt{All} is also kind polymorphic. Here, it applies the constraint to each inner list of types in the the outer list of lists.}. Using constraints as types of a special kind \texttt{Constraint}, hence they can appear as type parameters, is allowed by the \textsf{ConstraintKinds} language extension.

The \texttt{Generic} class definition in the \textsf{generics-sop} library allows to create the instance for a particular datatype automatically using internal \textsf{GHC.Generics}.

@@

In the previous section, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances. We are now going to improve the solution and remove overlap.

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. Equations for the \emph{closed type family} are disallowed to de defined outside its declaration. Under this extension, we can give the following definition of type-level equality:
\texttt{
\vs
\indent\textbf{type family} Equal a x \ann Bool \textbf{where}\\
\indent\s\s Equal a a = 'True\\
\indent\s\s Equal a x = 'False
\vs
}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is always strictly defined, overlapping equations can be used to define sound type-level functions like this type equality.

We now return to our running example from Section~\ref{sec:recursion-problem}. With type equality, we can witness the coersion between the equal types by defining a type class:
\texttt{
\vs
\indent\textbf{class} Proof (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s witnessEq \ann b \ra Maybe a
\vs
\indent\textbf{instance} Proof 'False a b \textbf{where}\\
\indent\s\s witnessEq \_ \hs\nths= Nothing\\
\indent\textbf{instance} Proof 'True\s\s a a \textbf{where}\\
\indent\s\s witnessEq\s\s\s = Just
\vs
}
We then need to provide the proof for all elements in \texttt{NP}. This can be done by means of the \texttt{All} combinator. For using that, we need the constraint can be partially applied, and so use a synonym for \texttt{Proof (Equal a b) a b} that we define as a class with a superclass constraint, rather than as a type synonym:
\texttt{
\vs
\indent\textbf{class}\ind Proof (Equal a b) a b \Ra ProofEq a b\\
\indent\textbf{instance} Proof (Equal a b) a b \Ra ProofEq a b
\vs
}
The following implementation of \texttt{subtermsNP} using this proof resembles our first definition given in the previous section:
\texttt{
\vs
\indent subtermsNP \ann \fa a xs. All (ProofEq a) xs \Ra NP I xs \ra [a]\\
\indent subtermsNP (I (y \ann x) :* ys)\\
\indent\s\s = case witnessEq @(Equal a x) y of\\
\indent\ind\s\s Just t\s\s \ra t : subtermsNP ys\\
\indent\ind\s\s  Nothing \ra subtermsNP  ys\\
\indent subtermsNP \_ Nil = []
\vs
}
We also make use of \textsf{ScopedTypeVariables} in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list. A proper \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.

All generic functions accessing recursive knots in the underlying datatype structure can be defined in this way. We give another detailed example of one of such functions to show how to adapt our idiom to different scenarios.

>>>>

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to represent a datatype as an isomorphic code whose kind is a list of lists of types. The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
\texttt{
\vs
\indent\textbf{type family} Code (a \ann *) \ann [[*]]
\vs
}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.

Consider, for instance, a datatype of binary trees:
\texttt{
\vs
\indent\textbf{data} Tree a = Leaf a | Node (Tree a) (Tree a)
\vs
}
This datatype is isomorphic to the following code:
\texttt{
\vs
\indent\textbf{type instance} Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\vspace{0.2cm}
}

@@

%A type of promoted lists has no inhabitants, so the universe provides datatypes to operate on $n$-ary sums and products as on terms.
% -- this sentence falls out of the flow of the paragraph

As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types. The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the code. Therefore, \texttt{NP} is a tiny abstraction of a usual definition of a heterogeneous list The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic. The index list allows to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.

Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analog of \texttt{const}:
\texttt{
\vs
\indent\textbf{newtype} I (a \ann *)\ind\ind\s\s\nhs = I \{unI \ann a\}\\
\indent\textbf{newtype} K (a \ann *) (b \ann k) = K \{unK \ann a\}
\vs
}

If instantiated with \texttt{I}, \texttt{NP} is a plain heterogeneous list of types, while \texttt{K a} turns it into homogeneous one. Here is an example value of \texttt{NP I}:
\texttt{
\vs
\indent I 5 :* I True :* I 'x' :* Nil \ann NP I '[Int, Bool, Char]
\vspace{0.2cm}
}

We turn to the sum definition now. The constructor \texttt{S} of \texttt{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \texttt{Z} stores the payload of type \texttt{f x}. For example, the following chooses the third element of a sum:
\texttt{
\vs
\indent S (S (Z (I 3))) \ann NS I '[Char, Bool, Int, Bool]
\vs
}
The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.

With the \texttt{NS} and \texttt{NP} machinery at hand, SOP defines the \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} witnessing the isomorphism between a datatype and its generic representation:
\texttt{
\vs
\indent\textbf{type} Rep a = SOP I (Code a)
\vs
\indent\textbf{class} All SListI (Code a) \Ra Generic (a \ann *) \textbf{where}\\
\indent\s\s \textbf{type} Code a \ann [[*]]\\
\indent\s\s from \ann a \ra Rep a\\
\indent\s\s to\s\s\s \ann Rep a \ra a
\vs
}
The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code. The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.

We are leaving out discussion of the \texttt{All SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work. Although, we do use \texttt{All} constraint combinator in the following. Therefore, it is worth noting that \texttt{All} applies particular constraint (e.g. \texttt{SListI} above) to each member of a list of types. The usage of constraints as type arguments allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.

We have introduced generic representation employed by the SOP library and are ready to describe the problem of handling recursion points stemming from the representation.

@@

In the previous section, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances. We are going to improve the solution and remove overlap now.

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. The main idea of the extension being that he equations for a \emph{closed type family} are disallowed outside its declaration. Under the extension, we can give the following definition of a type-level equality:
\texttt{
\vs
\indent\textbf{type family} Equal a x \ann Bool \textbf{where}\\
\indent\s\s Equal a a = 'True\\
\indent\s\s Equal a x = 'False
\vs
}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.

\subsection{Solution to \texttt{subterms} revised}

We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between the equal types by defining a type class:
\texttt{
\vs
\indent\textbf{class} Proof (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s witnessEq \ann b \ra Maybe a
\vs
\indent\textbf{instance} Proof 'False a b \textbf{where}\\
\indent\s\s witnessEq \_ \hs\nths= Nothing\\
\indent\textbf{instance} Proof 'True\s\s a a \textbf{where}\\
\indent\s\s witnessEq\s\s\s = Just
\vs
}

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \texttt{All} combinator and partially applied auxiliary type class \texttt{ProofEq} which abbreviates heavy-weighted interface of \texttt{Proof}:
\texttt{
\vs
\indent\textbf{class}\ind Proof (Equal a b) a b \Ra ProofEq a b\\
\indent\textbf{instance} Proof (Equal a b) a b \Ra ProofEq a b
\vs
}
Resulting implementation of \texttt{subtermsNP} resembles our first definition given in the previous section:
\texttt{
\vs
\indent subtermsNP \ann \fa a xs. All (ProofEq a) xs \Ra NP I xs \ra [a]\\
\indent subtermsNP (I (y \ann x) :* ys)\\
\indent\s\s = case witnessEq @(Equal a x) y of\\
\indent\ind\s\s Just t\s\s \ra t : subtermsNP ys\\
\indent\ind\s\s  Nothing \ra subtermsNP  ys\\
\indent subtermsNP \_ Nil = []
\vs
}

As a side note, we make use of \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.

To complete the solution of the problem, the \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms} task. We give another example showing how to adapt our idiom to different scenarios in the following section.
